---
title: "Activity: Bird Species"
author: "Allison Rachor"
date: "6/23/23"
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages
library(tidyverse)
library(dcData)



# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package
data("OrdwayBirds")



# Show a few commands used to inspect the data set a few different ways
head(OrdwayBirds)
tail(OrdwayBirds)
summary(OrdwayBirds)


```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds <-
  OrdwayBirds %>%
  select( SpeciesName, Month, Day ) %>%
  mutate( Month = as.numeric(as.character(Month)), 
          Day = as.numeric(as.character(Day)))
Corrected <-
  OrdwayBirds %>%
  inner_join( OrdwaySpeciesNames ) %>% 
  select( Species = SpeciesNameCleaned, Month, Day ) %>%
  na.omit()  # cleaned up the missing ones

# Inspect resulting data
head(OrdwayBirds)
head(Corrected)
```



### Step 1

The textbook points out the data set OrdwaySpeciesNames, which includes a variable called SpeciesNameCleaned. Summarizing both the original data set and the clean list, there are 166 more "unique" names in the original data set due to spelling mistakes (there were 275 unique names in OrdwayBirds and 109 in OrdwaySpeciesNames. 


```{r}
# unique species in original data (including mis-spellings)
OrdwayBirds %>%
  summarize(n_distinct(SpeciesName))



# unique species in the clean list
OrdwaySpeciesNames %>%
  summarize(n_distinct(SpeciesNameCleaned))

```


### Step 2
  
Joining the two data sets created a table with the correct spelling of all the species names, along with two variables that record the month and the day of sighting. The variables Species in OrwayBirds and the variable SpeciesNameCleaned were used to join the data sets. The variable recording the names in the original data set changed to only include correct spelling.


```{r}
# inspect OrdwaySpeciesNames
data("OrdwaySpeciesNames")
head(OrdwaySpeciesNames)
tail(OrdwaySpeciesNames)

# inner join
Corrected <-
  OrdwayBirds %>%
  inner_join( OrdwaySpeciesNames ) %>% 
  select( Species = SpeciesNameCleaned, Month, Day ) %>%
  na.omit()  # cleaned up the missing ones



# inspect the data after the join
head(Corrected)
summary(Corrected)

```


### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

The Corrected data set was used to group the species names together to allow for a total of times each species was sighted could be counted. A threshold of 700 was set so that only the top 6 sighted birds would be considered a major species in the area.

```{r}
# count sightings by species in descending order
Corrected %>%
  group_by(Species) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# a hint in the text book recommends displaying the top 10 species to choose our threshold

Corrected %>%
  group_by(Species) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10)

# define a sightings threshold; produce a table with only birds belonging to major species
threshold = 700

Majors <- Corrected %>%
  group_by(Species) %>%
  summarize(count = n()) %>%
  filter(count >= threshold)
head(Majors)



```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`

Majors2 <- c("American Goldfinch", "Black-capped Chickadee", "Field Sparrow", "Lincoln's Sparrow", "Slate-colored Junco")

ByMonth <- Corrected %>% 
  filter(Species %in% Majors2) %>%
  group_by(Species, Month) %>%
  summarize(count = sum(Month))





```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = count, color = Species)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=count ,fill=Species)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  

The only species seen year-round is the Black-capped Chickadee.

##### 2. Which species are migratory, that is, primarily present in one or two seasons?

Slate-colored Junco an the Lincoln's Sparrow look like the only birds that are present in only 2 seasons.


##### 3. What is the peak month for each major species? (bullet list)
American Goldfinch- October
Black-capped Chickadee - September
Field Sparrow - October
Lincoln's Sparrow - May
Slate-colored Junco - October



##### 4. Which major species are seen in good numbers for at least 6 months of the year?

The American Goldfinch is the only species that is seen in decent numbers for at least 6 months.

```{r}

```

